---
title: "The Post-2007 Declining U.S. Birth Rates imply Significant Societal Shifts in the U.S."
author: 
  - Sagith Kalaichelvam
  - Youngho Kim
  - Joseph Chung
thanks: "Code and data are available at: https://github.com/UtopianYoungChung/Paper2.git"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper explores several factors of the steep decline in birth rates in the United States between 2007 and 2020. The decline in U.S. birth rates offers some significant insights into U.S. society. Changes in birth patterns among recent young adult women moving through the childbearing age group can be directly related to the decline in birth rate from a purely statistical standpoint. In this paper, we decompose the decline in birth rates into between - and within - demographic groups. We then argue that any other economic, policy or social factor that has changed since 2007 does not contribute much to the decline. We also provide evidence that the steep drop in birth rates could be attributed to recent young adults' shifting interests. These shifting priorities represent changes in young adults' preferences regarding having children and aspirations for life. Resonating with the original paper, we conclude with a brief discussion of the implications of the declining birth rate in the United States and its impact on society as an entirety."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(knitr)
library(ggplot2)
```

# Introduction

We need Figure 1, Figure 2 - A & B, Table 1, Figure 4, and Figure 5 to show the evidence for our claim.

You can and should cross-reference sections and sub-sections. For instance, @sec-data and @sec-first-point.

# Data {#sec-data}

Figure 1. place it here with the Figure 1 from the Original paper (@fig-bills).

```{r}
fig1_data <- read.csv("data/fig_1.csv")
fig1_data %>%
  ggplot(aes(x = year,
             y = brate_all)) +
  geom_line() +
  scale_y_continuous(limits = c(50, 80), breaks = seq(50, 80, by = 5)) +
  labs(
    title = "Trend in US Birth Rates",
    x = "Years",
    y = "Births per 1,000 women age 15-44",
    caption = "Source: Birth Rates collected from CDC Vital Statistics Births Reports for 2015, 2019 and 2020."
  )  +
  theme_light()
```

Talk more about it.

Table 1. place it here with the table 1 from the Original paper

```{r}


```

Figure 2. place it here with the Figure 4 from the Original paper (@fig-billss).

```{r}
#| label: fig-billss
#| fig-cap: Relationship between Changes in Birth Rates 2004-2008 to 2014-2019
#| echo: false
#| fig-width: 8
#| fig-height: 4

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Figure 3. place it here with the Figure 2-a from the Original paper
```{r}
#| label: fig-billsss
#| fig-cap: Trends in Birth Rates by Population subgroup Five-year age group
#| echo: false
#| fig-width: 8
#| fig-height: 4

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Figure 4. place it here with the Figure 5 from the Original paper
```{r}
#| label: fig-billssss
#| fig-cap: Children Ever Born by Mother's Age and Birth Cohort
#| echo: false
#| fig-width: 8
#| fig-height: 4

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()

```

# Results

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional details

\newpage

# References
